---
title: "Coral Demography Raw Data Cleaning"
format: html
editor: visual
authors: Kylie Newcomer, Joaquin Sandoval, Vedika Shirtekar
---

This document handles cleaning and quality assurance of long-format coral demography data spanning 2013–2023, as well as the standardization of 2024 field sheets to align with the long-format structure. Each row represents a single coral observation in a given year, with associated size measurements (length, width, height) and demographic dynamics (recruitment, death, fusion, fission etc). The cleaning steps standardize inconsistent labels in the `Dynamics` column, split it into binary indicator columns, and address logical inconsistencies in recruitment and death events to ensure each represents a true demographic transition.

### Load Necessary Libraries

```{r}
#| message: false

library(tidyverse)
library(readxl)
library(here)
library(janitor)
```

### Import `Coral_Deomography_allyears_2013-24_V1.xlsx` and 2024 raw field sheets

```{r}
#| message: false
#| warning: false

# Sheet 1 contains 2013–2023 only; 2024 comes from field_sheets_2024
wide_coral_demography_2013_2023 <- read_excel(here("rawdata", "Coral_Deomography_allyears_2013-24_V1.xlsx"), sheet = 1)

# Produce a character vector of the names of files in 2024 sheets folder 
field_sheets_2024 <- list.files(here("rawdata", "2024sheets"), pattern = "\\.xlsx$", full.names = TRUE)

```

### Combine 2024 field sheets into a single dataframe

```{r}
#| message: false

# One-time cleanup function for raw 2024 sheets with inconsistent column names.
# Pulls observer name from the last non-NA cell of row 1 (layout varies across sheets).
# Future sheets will be standardized and won't need this.
read_file <- function(f) {
  meta <- read_excel(f, col_names = FALSE, n_max = 1)
  obs <- tail(na.omit(as.character(meta[1, ])), 1)
  read_excel(f, skip = 1, col_names = TRUE, col_types = "text") %>%
    mutate(Observer = obs)
}

# Combine into one dataframe
coral_demography_2024 <- bind_rows(lapply(field_sheets_2024, read_file)) %>%
  # Drop columns with previous year measurements
  select(-c("W", "H", "L", "Status")) %>%
  # Correct naming convention
  rename(habitat = Hab,
         transect = Tran) %>%
  # Standardize column naming structure across sheets
  mutate(L2024 = coalesce(L2024, L24),
         W2024 = coalesce(W2024, W24),
         H2024 = coalesce(H2024, H24),
         # Standardize dynamics column across sheets
         Dynamics_2024 = coalesce(Dynam, Dynamic, Dyn, Dynamics),
         across(c(X, Y, Z), as.numeric)) %>%
  select(-any_of(c("L24", "W24", "H24",
                   "Dynam", "Dynamic", "Dyn", "Dynamics", "...17",
                   "status", "source_file"))) %>%
  #"A" was used for Acropora in 2024; standardize to "Acr"
  mutate(Taxa   = if_else(Taxa == "A", "Acr", Taxa)) %>% 
  clean_names()

# Verify only 4 taxa exist 
cat("Unique taxa:", sort(unique(na.omit(coral_demography_2024$taxa))))
```

### Remove empty 2024 columns from `wide_coral_demography_2013_2023` and full join with 2024 observations `coral_demography_2024`

```{r}
wide_coral_demography_2013_2023 <- wide_coral_demography_2013_2023 %>%
  select(!contains("24")) %>% # Remove 2024 columns because they have no data
  clean_names()

wide_coral_demography_2013_2024 <- full_join(wide_coral_demography_2013_2023, coral_demography_2024, 
                        by = c("site", "habitat", "transect", "taxa", "x", "y", "z")) %>% 
  # Remove any "filler" or empty rows
  filter(!is.na(site))
```

### Creating new coral number for 2024 recruits

```{r}
# Assign new coral numbers to 2024 recruits (rows with NA), starting from max + 1
wide_coral_demography_2013_2024 <- wide_coral_demography_2013_2024 %>%
  mutate(coral_number = if_else(
    is.na(coral_number),
    max(coral_number, na.rm = TRUE) + cumsum(is.na(coral_number)),
    coral_number
  ))
# Verify all coral numbers are unique
if (length(unique(wide_coral_demography_2013_2024$coral_number)) ==
    nrow(wide_coral_demography_2013_2024)) {
  print("All coral numbers are unique")
} else {
  print("Duplicate coral numbers found")
}
```

### Pivot complete dataset longer

Each row represents a single coral observation in a given year. We found several instances of coral numbers with no measurement data.

```{r}
# Pivot year-suffix columns to long format (one row per coral-year)
long_coral_demography_2013_2024 <- wide_coral_demography_2013_2024 %>%
  pivot_longer(cols = -c(1:8), names_to = c(".value", "year"), names_pattern = "(.*)(\\d{4}$)") %>% 
  # Filter out empty rows from 2024 sheets 
  filter(!is.na(year))
```

### Expand and standardize dynamics column

```{r}
#| warning: false

# Separate cells with multiple dynamics into multiple rows, then pivot wide
long_coral_demography_2013_2024_dynamics <- long_coral_demography_2013_2024 %>%
  separate_longer_delim(dynamics_, delim = regex("\\s*,\\s*")) %>% # regex for inconsistencies in spacing
  mutate(present = case_when(
    is.na(dynamics_) ~ 0, # If dynamics column is NA, present = 0
    TRUE ~ 1),
    # Standardize the dynamics_ labels
    dynamics_ = case_when(
      str_detect(dynamics_, "Fus") ~ "Fusion",
      str_detect(dynamics_, "Fis") ~ "Fission",
      dynamics_ == "Dead"          ~ "Death",
      TRUE ~ dynamics_
    )) %>%
  pivot_wider(names_from = dynamics_, values_from = present, names_prefix = "dyn_", values_fill = 0) %>%
  clean_names()
```

### Correct transect assignments for misidentified corals

```{r}
long_coral_demography_2013_2024_dynamics <- long_coral_demography_2013_2024_dynamics %>%
  mutate(transect = case_when(coral_number %in% c(1347: 1353) ~ "T01",
                              coral_number %in% c(1361: 1363) ~ "T01",
                              coral_number %in% c(1366: 1376) ~ "T01",
                              coral_number %in% c(1466: 1483) ~ "T01",
                              coral_number == 1863 ~ "T04",
                              coral_number == 1872 ~ "T04",
                              TRUE ~ transect))
```

### Coerce measurement columns to numeric and select final columns

```{r}
#| warning: false
long_coral_demography_2013_2024_dynamics <- long_coral_demography_2013_2024_dynamics %>%
  mutate(across(c(year, l, w, h), as.numeric)) %>%
  select(coral_number, year, site, habitat, transect, taxa, x, y, z, l, w, h, obs, note,
         dyn_recruitment, dyn_growth, dyn_shrinkage, dyn_death, dyn_immigration, dyn_emigration, dyn_fusion,
         dyn_fission, dyn_edge)
```

```{r}
# Filtering out corals with NAs across all measurements 
long_coral_demography_2013_2024_dynamics <- long_coral_demography_2013_2024_dynamics %>% 
  group_by(coral_number) %>% 
  mutate(lengths_sum = sum(l, na.rm = TRUE)) %>% # Take sum of lengths across years
  filter(lengths_sum > 0) # Select only corals with at least one length measurement
```

### Quality Assurance for dynamic columns

```{r}
# A coral should only be recruited once — flag any with >1 recruitment event
recruit_check <- long_coral_demography_2013_2024_dynamics %>%
  group_by(coral_number) %>%
  summarize(recruit_occurrences = sum(dyn_recruitment)) %>%
  filter(recruit_occurrences > 1)

if (nrow(recruit_check) > 0) {
  warning(nrow(recruit_check), " corals have multiple recruitment events — see recruit_check")
}
```

### Remove duplicate recruitment events

```{r}
recruit_fix <- long_coral_demography_2013_2024_dynamics %>%
  arrange(coral_number, year) %>%
  group_by(coral_number) %>%
  # Keep only the first recruitment event per coral
  mutate(dyn_recruitment = if_else(cumsum(dyn_recruitment) > 1, 0, dyn_recruitment)) %>%
  ungroup()
```

### Verify duplicate recruitments are resolved

```{r}
recruit_fix_check <- recruit_fix %>%
  group_by(coral_number) %>%
  summarize(recruit_occurrences = sum(dyn_recruitment)) %>%
  filter(recruit_occurrences > 1)

if (nrow(recruit_fix_check) == 0) {
  message("All corals now have at most one recruitment event.")
} else {
  warning(nrow(recruit_fix_check), " corals still have multiple recruitment events.")
}
```

### Quality assurance for death events

```{r}
# A coral should only die once — flag any with >1 death event
death_check <- recruit_fix %>%
  group_by(coral_number) %>%
  summarize(total_death = sum(dyn_death)) %>%
  filter(total_death > 1)

if (nrow(death_check) == 0) {
  message("All corals have at most one death event.")
} else {
  warning(nrow(death_check), " corals have multiple death events — see multi_death_histories")
}
```

### Inspect histories of corals with multiple death events

```{r}
# Pull full year-by-year history for flagged corals to review manually
multi_death_histories <- recruit_fix %>%
  filter(coral_number %in% death_check$coral_number) %>%
  arrange(coral_number, year) %>%
  select(coral_number, year, l, w, h, dyn_recruitment, dyn_death)
```

### Remove duplicate death events

-- show instances of corals having measurements, dying and then at most 2 years later, having measurements again

```{r}
death_fix <- recruit_fix %>%
  arrange(coral_number, year) %>%
  group_by(coral_number) %>%
  # Correct any corals that "died" but continued to have measurements within 2 years because this was longest gap of missing data between 1st and measurements appearing 
  mutate(dyn_death = ifelse(dyn_death == 1 & lead(!is.na(l), n = 2) & year < 2022, 0, dyn_death)) %>%
  mutate(
    # Create binary is measurements are recorded
   has_measurements = !is.na(l),
    # Track whether coral has died at any point before
    died_before = lag(cumsum(dyn_death), default = 0) >= 1,
    # New life detected: measurements present AFTER a previous death
    new_life = has_measurements & died_before,
    # Reset the "dead" status when a new life starts
    life_stage = cumsum(new_life) + 1,
    # Create a new coral number for new corals
    new_coral_number = case_when(
      # Append life stage so each reincarnation gets a unique ID
      life_stage > 1 ~ paste0(coral_number, "_", life_stage),
      TRUE ~ as.character(coral_number)
    )
  ) %>%
  # Group by NEW coral number
  group_by(new_coral_number) %>%
  # Within each life, keep only the first death
  mutate(dyn_death = if_else(cumsum(dyn_death) > 1, 0, dyn_death)) %>%
  ungroup() %>%
  select(-c(has_measurements, died_before, new_life))
```

### Verify duplicate deaths are resolved

```{r}
death_fix_check <- death_fix %>%
  group_by(new_coral_number) %>%
  summarize(total_death = sum(dyn_death)) %>%
  filter(total_death > 1)

if (nrow(death_fix_check) == 0) {
  message("All corals have at most one death event.")
} else {
  warning(nrow(death_fix_check), " corals have multiple death events – see multi_death_histories")
}
```

### Assign coral_id and transect_id

```{r}
### Assign coral_id and transect_id
corals_long_with_coral_id <- death_fix %>%
arrange(new_coral_number, year) %>% 
          group_by(new_coral_number) %>%
# New variable signifying when a coral has measurements
mutate(years_measure = ifelse(!is.na(l), year, NA)) %>%
# Select the first year of measurements
mutate(first_year = min(years_measure, na.rm = TRUE),
# Create new coral_id based on location, taxa, and first occurance
coral_id = str_c(site, habitat, transect, taxa, x, y, z, first_year, sep = "-"),
transect_id = str_c(site, habitat, transect, sep = "_"))  %>% 
  ungroup() %>% 
select(c(year, coral_id, site, habitat, transect, transect_id, taxa, x, y, z, l, w, h, obs, note, dyn_recruitment, dyn_growth, dyn_shrinkage, dyn_death, dyn_immigration, dyn_emigration, dyn_fusion, dyn_fission, dyn_edge))
```

### Quality assurance for coral positions

Verify all transect positions are within transect boundaries

x dimension: moving forward on the transect - meters (0-5) y dimension: left & right from x vector - centimeters (0-100) z dimension: vertical position from benthos - centimeters

Transects: One 5x1m transect on top of each of four 5x5m plot

Plots: Ten 10x1m transects running through 10x10m plot

```{r}
# Transect QA - x positions out of bounds
transects <- corals_long_with_coral_id %>% filter(str_detect(transect, "^T"))
t_oob_x <- transects %>% filter(x < 0 | x > 5) %>% distinct(coral_id)
cat("Transects - corals with x coordinate out of bounds:", nrow(t_oob_x), "\n")

# Plot QA - x positions out of bounds
plots <- corals_long_with_coral_id %>% filter(str_detect(transect, "^P"))
p_oob_x <- plots %>% filter(x < 0 | x > 10) %>% distinct(coral_id)
cat("Plots - corals with x coordinate out of bounds:", nrow(p_oob_x), "\n")
```

## Fix x values out of bounds

```{r}
# Check all out-of-bounds x values before fixing
corals_long_with_coral_id %>% 
  filter((str_detect(transect, "^T") & x > 5) | 
         (str_detect(transect, "^P") & x > 10)) %>% 
  select(coral_id, transect, taxa, x) %>% 
  distinct()
```

```{r}
# Clear factor-of-10 errors (x >> boundary)
corals_long_with_coral_id <- corals_long_with_coral_id %>%
  mutate(x = if_else(str_detect(transect, "^T") & x > 10, x / 10, x),
         x = if_else(str_detect(transect, "^P") & x > 20, x / 10, x))

# Transect QA - positions out of bounds
transects <- corals_long_with_coral_id %>% filter(str_detect(transect, "^T"))
t_oob_x <- transects %>% filter(x < 0 | x > 5) %>% distinct(coral_id)
cat("Transects - corals with x coordinate out of bounds:", nrow(t_oob_x), "\n")

# Plot QA - positions out of bounds
plots <- corals_long_with_coral_id %>% filter(str_detect(transect, "^P"))
p_oob_x <- plots %>% filter(x < 0 | x > 10) %>% distinct(coral_id)
cat("Plots - corals with x coordinate out of bounds:", nrow(p_oob_x), "\n")
```

```{r}
# Borderline cases - flag for manual review
corals_long_with_coral_id %>%
  filter((str_detect(transect, "^T") & x > 5) |
         (str_detect(transect, "^P") & x > 10)) %>%
  select(coral_id, transect, taxa, x) %>%
  distinct()
```

## Correct borderline cases 

```{r}
corals_long_with_coral_id <- corals_long_with_coral_id %>%
  mutate(x = case_when(
    coral_id == "LTER2-OR-T04-Acr-5.4-39-5-2019" ~ 0.54,
    coral_id == "LTER4-BR-T01-Poc-8-50-30-2020"  ~ 0.8,
    coral_id == "LTER4-OR-T03-Por-8-75-0-2020"   ~ 0.8,
    coral_id == "LTER5-OR-T02-Acr-9.5-17-3.4-2018" ~ 0.95,
    coral_id == "LTER1-BR-T03-Poc-5.1-90-90-2024" ~ 0.51,
    coral_id == "LTER1-BR-T03-Poc-6-4.5-30-2024" ~ 0.60,
    TRUE ~ x
  ))
```

```{r}
# Transect QA - verfity no x positions out of bounds
transects <- corals_long_with_coral_id %>% filter(str_detect(transect, "^T"))
t_oob_x <- transects %>% filter(x < 0 | x > 5) %>% distinct(coral_id)
cat("Transects - corals with x coordinate out of bounds:", nrow(t_oob_x), "\n")
```

## Fix y values out of bounds

```{r}
# Transect QA - y positions out of bounds
transects <- corals_long_with_coral_id %>% filter(str_detect(transect, "^T"))
t_oob_y <- transects %>% filter(y < 0 | y > 100) %>% distinct(coral_id)
cat("Transects - corals with y coordinate out of bounds:", nrow(t_oob_y), "\n")

# Plot QA - y positions out of bounds
plots <- corals_long_with_coral_id %>% filter(str_detect(transect, "^P"))
p_oob_y <- plots %>% filter(y < 0 | y > 100) %>% distinct(coral_id)
cat("Plots - corals with y coordinate out of bounds:", nrow(p_oob_y), "\n")
```

```{r}
# Check individual coral with out-of-bounds y value before fixing
corals_long_with_coral_id %>% 
  filter((str_detect(transect, "^P") & y > 100)) %>% 
  select(coral_id, transect, taxa, y) %>% 
  distinct()
```

```{r}
# Look at history. All y values are 200
corals_long_with_coral_id %>% 
  filter(coral_id == "LTER4-BR-P07-Acr-2.25-200-0-2021") %>% 
  select(year, coral_id, x, y, z)
```

```{r}
# Correct to 20 cm on transect 
corals_long_with_coral_id <- corals_long_with_coral_id |>
  mutate(
    y = if_else(coral_id == "LTER4-BR-P07-Acr-2.25-200-0-2021", 20, y)
  )
```

```{r}
# Plot QA - y positions out of bounds
plots <- corals_long_with_coral_id %>% filter(str_detect(transect, "^P"))
p_oob_y <- plots %>% filter(y < 0 | y > 100) %>% distinct(coral_id)
cat("Plots - corals with y coordinate out of bounds:", nrow(p_oob_y), "\n")
```

### Quality assurance for coral measurements

Verify all minimum/maximum measurements are reasonable.

```{r}
# Range of each measurement for each taxa/habitat combination 
corals_long_with_coral_id %>%
  group_by(taxa, habitat) %>%
  summarise(across(c(l, w, h), list(min = ~min(., na.rm = TRUE), 
                                      max = ~max(., na.rm = TRUE))), 
            .groups = "drop_last")    # Silences message 
```

### Flag Corals with large measurements for individual review.

Rough estimates of the maximum measurements of all genera were given by Dr. Hunter Lenihan as follows:

*Pocillopora*: \~1 meter

*Acropora*: \~2 meters

*Porites*: \~4 meters

*Millepora*: \~2 meters

```{r}
corals_long_with_coral_id |>
  mutate(
    max_dim = case_when(
      taxa == "Poc" ~ 100,
      taxa == "Acr" ~ 200,
      taxa == "Por" ~ 400,
      taxa == "Mil" ~ 200
    )
  ) |>
  filter(if_any(c(l, w, h), ~ .x > max_dim)) |>
  select(year, coral_id, site, habitat, transect, taxa, l, w, h)
```

```{r}
# View life history of flagged Pocillopora 
corals_long_with_coral_id %>% 
  filter(coral_id == "LTER2-BR-P04-Poc-1.45-55-40-2023") %>% 
  select(year, coral_id, l, w, h)
```

```{r}
# Correct 2024 measurements 
corals_long_with_coral_id <- corals_long_with_coral_id |>
  mutate(
    l = if_else(coral_id == "LTER2-BR-P04-Poc-1.45-55-40-2023" & year == 2024, 2.52, l),
    w = if_else(coral_id == "LTER2-BR-P04-Poc-1.45-55-40-2023" & year == 2024, 4, w),
    h = if_else(coral_id == "LTER2-BR-P04-Poc-1.45-55-40-2023" & year == 2024, 4, h)
  )
```

```{r}
# View life history of flagged Porites 
corals_long_with_coral_id %>% 
  filter(coral_id == "LTER4-BR-T04-Por-5-30-40-2013") %>% 
  select(year, coral_id, l, w, h)
```

```{r}
# Correcting incorrect height of Porites individual in 2020 
corals_long_with_coral_id <- corals_long_with_coral_id |>
  mutate(
    h = if_else(coral_id == "LTER4-BR-T04-Por-5-30-40-2013" & year == 2020, 80, h)
  )
```

### Minimum measurements are mostly accurate

There is one individual with 0 as as a height measurement.

```{r}
# Isolate coral for coral_id
corals_long_with_coral_id %>% 
  filter(h == 0) %>% 
  select(year, coral_id, l, w, h)

# View life history 
corals_long_with_coral_id %>% 
  filter(coral_id == "LTER1-BR-P02-Acr-6.3-30-2-2023") %>% 
  select(year, coral_id, l, w, h)
```

```{r}
# Correct measurement 
corals_long_with_coral_id <- corals_long_with_coral_id |>
  mutate(
    h = if_else(coral_id == "LTER1-BR-P02-Acr-6.3-30-2-2023" & year == 2023, 10, h)
  )
```

```{r}
# Recheck range of each measurement for each taxa/habitat combination 
corals_long_with_coral_id %>%
  group_by(taxa, habitat) %>%
  summarise(across(c(l, w, h), list(min = ~min(., na.rm = TRUE), 
                                      max = ~max(., na.rm = TRUE))), 
            .groups = "drop_last")    # Silences message 
```

### Write .csv to `projectdata` folder.

`coral_demography_2013_2024_long.csv` will be loaded into `coral_demography_analysis_cleaning.qmd` for transformation into an analysis-ready dataset used by downstream GLMMs, demographic profile figures, and spatiotemporal visualizations.

```{r}
 write_csv(corals_long_with_coral_id,  here::here("projectdata", "coral_demography_2013_2024_long.csv"))
```
