---
title: "Coral Demography Analysis Cleaning Script "
author: Kylie Newcomer, Joaquin Sandoval, Vedika Shirtekar 
date: 4/19/2026
format: html
editor: visual
---

This script prepares `coral_demography_2013_2024_long.csv` for downstream GLMM analysis, profile figures, and transect-level spatiotemporal visualizations. The input file is read from the output of the `coral_demography_raw_data_cleaning.qmd` : `coral_demography_2013_2024_long.csv`.

**Data Cleaning**

- Filters to transect data only, excluding backreef plots

- Filters for only recruitment and death dynamic columns

- Converts `y` and `z` dimensions from meters to centimeters for downstream visualizations

**Variable Construction**

- Creates a bleaching period indicator: `0` = prior (2013–2018), `1` = post (2019–2024)

- Sources volume formulas from the `R/` folder to calculate colony volume by taxa: cylinder (*Acropora*), hemisphere *Porites*), sphere (*Pocillopora*)

- Creates lagged size variables: `size_t_minus1` (previous year), `size_t0` (current year), `size_t1` (following year)

### Load Necessary Libraries

```{r}
#| message: false
library(tidyverse)
library(here)
```

### Read in `coral_demography_2013_2024_long.csv`

```{r}
#| message: false
coral_demography_2013_2024_long <- read_csv(here("projectdata", "coral_demography_2013_2024_long.csv"))
```

# Source volume functions

```{r}
source(here::here("R", "sph_vol.R"))

source(here::here("R", "hemisph_vol.R"))

source(here::here("R", "cyl_vol.R"))
```

# Calculate volume and lagged sizes

Removed rowwise() and the 3-way filter/rbind — replaced with a single case_when in mutate Dropped size_t0 = volume (redundant) Added ungroup() after the lagged size calculation Switched case_when to if_else for the single taxa recode here::here() on both write_csv calls Collapsed the initial cleaning into one piped chain

```{r}
coral_demography_2013_2024_analysis <- coral_demography_2013_2024_long %>%
  mutate(size_t0 = case_when(
    taxa == "Poc" ~ sph_vol(l, w, h),
    taxa == "Por" ~ hemisph_vol(l, w, h),
    TRUE          ~ cyl_vol(l, w, h)
  )) %>%
  arrange(coral_id, year) %>%
  group_by(coral_id) %>%
  mutate(
    size_t_minus1 = lag(size_t0),
    size_t1       = lead(size_t0)
  ) %>%
  ungroup() %>% 
  filter(str_starts(transect, "T")) %>% 
  select(year, coral_id, site, habitat, transect, transect_id, taxa, x, y, z, l, w, h, size_t_minus1, size_t0, size_t1 ,obs, note, dyn_recruitment, dyn_death)
```

```{r}
write_csv(coral_demography_2013_2024_analysis, here::here("analysisdata", "coral_demography_analysis_data.csv"))
```
