---
title: "Coral Demography Volume Cleaning Script "
author: Kylie Newcomer, Joaquin Sandoval, Vedika Shirtekar 
date: 4/19/2026
format: html
editor: visual
---

This script further cleans, standardizes, and prepares `updated_coral_tidy_2013-2024.csv` for downstream GLMM analysis, profile figures, and transect-level spatiotemporal visualizations. The input file was obtained after initial raw data processing by Dr. Hunter Lenihand and Ryan Anderson.

**Data Cleaning**

-   Standardizes missing and ambiguous character values (`MD`, `UK`, `MK`, `?`, `MissingObs`, `Na`, `NA`, `Missing Data MD (Not sampled in 2017)`) to R `NA`

-   Converts measurement columns to numeric after character removal

-   Filters to transect data only, excluding backreef plots

-   Converts `y` and `z` dimensions from meters to centimeters for downstream visualizations

**Variable Construction**

-   Reassigns updated transect versions to a unified transect column

-   Creates a unique transect ID

-   Creates a bleaching period indicator: `0` = prior (2013–2018), `1` = post (2019–2024)

-   Sources volume formulas from the `R/` folder to calculate colony volume by taxa: cylinder (*Acropora*), hemisphere *Porites*), sphere (*Pocillopora*)

-   Creates lagged size variables: `size_t_minus1` (previous year), `size_t0` (current year), `size_t1` (following year)

**Coral Identity Validation**

-   Checks for duplicate `coral_id` usage across distinct individuals — if a coral dies and a new recruit settles at the same coordinates, the same `coral_id` may reappear after a gap

-   Appends the first observed year to `coral_id` (`site_habitat_transect_x_y_z_firstyear`) to uniquely identify individuals

-   If a coral with the same `coral_id` decreases in size by \>95% between consecutive years, it is split into two separate coral numbers, as this likely represents a new individual

```{r}
#| message : false

library(tidyverse)

coral_colors <- c(
 Acr = "blue", 
 Poc = "red", 
 Por = "darkgreen", 
 Mil = "darkorange"
)
```

```{r}
# Read in data, converting missing codes to NA
clean_coral <- read_csv(here::here("rawdata", "long_corals.csv")) %>%
  # Remove backreef plots, keep only transect surveys
  filter(str_detect(transect, "T")) %>%
 mutate(
    height = as.numeric(h),
    width  = as.numeric(w),
    length = as.numeric(l),
    # Fix data entry error: height 800 should be 80 for this porites in 2020
    height = if_else(coral_id == "LTER4_BR_T04_Por_5_30_40" & year == 2020, 80, height),
    y_m    = y / 100,
    z_m    = z / 100,
    # "A" was used for Acropora in 2024; standardize to "Acr"
    taxa   = if_else(taxa == "A", "Acr", taxa),
    # Assign old transects
    transect = case_when(
           str_detect(transect, "1") ~ "T1",
           str_detect(transect, "2") ~ "T2",
           str_detect(transect, "3") ~ "T3",
           str_detect(transect, "4") ~ "T4",
         ),
    # Create bleaching period variable
    period = case_when(
    year < 2019 ~ 0,
    TRUE ~ 1
  ),
  )
```

```{r}
## Finding giant porites 

# clean_coral %>% 
#   filter(taxa == "Por") %>% 
#   arrange(desc(volume)) %>% 
#   select(coral_id, year, length, width, height, volume) %>% 
#   head()
```

# Source volume functions

```{r}
source(here::here("R", "sph_vol.R"))

source(here::here("R", "hemisph_vol.R"))

source(here::here("R", "cyl_vol.R"))
```

# Calculate volume and lagged sizes

Removed rowwise() and the 3-way filter/rbind — replaced with a single case_when in mutate Dropped size_t0 = volume (redundant) Added ungroup() after the lagged size calculation Switched case_when to if_else for the single taxa recode here::here() on both write_csv calls Collapsed the initial cleaning into one piped chain

```{r}
volume_coral <- clean_coral %>%
  mutate(size_t0 = case_when(
    taxa == "Poc" ~ sph_vol(length, width, height),
    taxa == "Por" ~ hemisph_vol(length, width, height),
    TRUE          ~ cyl_vol(length, width, height)
  )) %>%
  arrange(coral_id, year) %>%
  group_by(coral_id) %>%
  mutate(
    size_t_minus1 = lag(size_t0),
    size_t1       = lead(size_t0)
  ) %>%
  ungroup()
```

For each coral, look at its years in order and check if there's ever a jump of more than 1 year between observations. If so, keep all rows for that coral. The result is a dataset of only the corals that "disappeared" for at least one survey year and reappeared later - this would indicate recruitment of the same species at the exact location where another had died.

```{r}
# For each coral_id, check if years are non-consecutive (gap = died & same species recruited)
gaps <- volume_coral |>
  arrange(coral_id, year) |>
  group_by(coral_id) |>
  filter(any(diff(year) > 1)) |> # any() Given set of logical vectors, is at least one value true? 
  ungroup()                      # diff() calculates difference between years 
  

# total number of instances where this took place 
n_distinct(gaps$coral_id)

```

```{r}
# Flag coral_ids where size drops dramatically (e.g., large coral -> tiny recruit)
resets <- volume_coral |>
  select(year, site, habitat, transect_id, coral_id, size_t0, size_t1) %>% 
  arrange(coral_id, year) |>
  group_by(coral_id) |>
  mutate(pct_change = (size_t1 - size_t0) / size_t0) |>
  filter(pct_change < -0.95) |>
  ungroup()

reset_ids <- resets |> pull(coral_id) |> unique()

resets <- volume_coral |>
  filter(coral_id %in% reset_ids) |>
  select(year, coral_id, size_t0, size_t1, dyn_death, dyn_recruitment) |>
  arrange(coral_id, year)
```

```{r}
write_csv(clean_coral,  here::here("analysisdata", "clean_corals.csv"))
write_csv(volume_coral, here::here("analysisdata", "volume_corals.csv"))
```

```{r}
ggplot(volume_coral, aes(x = taxa, y = size_t0, fill = taxa)) +
  geom_boxplot() +
  labs(x = "Taxa", y = "Log Volume (cm³)") +
  scale_y_log10(labels = scales::label_log()) + 
  scale_fill_manual(values = coral_colors) + 
  theme_minimal() + 
  theme(legend.position = "none")
```

```{r}
library(ggridges)


ggplot(volume_coral, aes(x = size_t0, y = taxa, fill = taxa)) +
  geom_density_ridges() +
  scale_x_log10(labels = scales::label_log()) +
  labs(x = "Log Volume (cm³)", y = "Taxa") +
  scale_fill_manual(values = coral_colors) + 
  theme_minimal() + 
    theme(legend.position = "none")
```
