---
title: "Volume Cleaning Script"
format: html
editor: visual
--- %>% %>% 
---

```{r}
library(tidyverse)

coral_colors <- c(
 Acr = "blue", 
 Poc = "red", 
 Por = "darkgreen", 
 Mil = "darkorange"
)
```

```{r}
# Read in data, converting missing codes to NA
clean_coral <- read_csv(
  here::here("data", "updated_coral_tidy_2013-2024.csv"),
  na = c("UK", "MK", "?", "MD", "US", "NS", "MissingObs", "Na", "NA")
) %>%
  # Remove backreef plots, keep only transect surveys
  filter(str_detect(transect, "T")) %>%
 mutate(
    height = as.numeric(height),
    width  = as.numeric(width),
    length = as.numeric(length),
    # Fix data entry error: height 800 should be 80 for this porites in 2020
    height = if_else(coral_number == "LTER4_BR_T04_Por_5_30_40" & year == 2020, 80, height),
    y_m    = y / 100,
    z_m    = z / 100,
    # "A" was used for Acropora in 2024; standardize to "Acr"
    taxa   = if_else(taxa == "A", "Acr", taxa),
    # Assign old transects
    transect = case_when(
           str_detect(transect, "1") ~ "T1",
           str_detect(transect, "2") ~ "T2",
           str_detect(transect, "3") ~ "T3",
           str_detect(transect, "4") ~ "T4",
         ),
    # Create unique transect ID
    transect_id = paste(site, habitat, transect, sep = "_"),
    # Create bleaching period variable
    period = case_when(
    year < 2019 ~ 0,
    TRUE ~ 1
  ),

  
  )
```

```{r}
## Finding giant porites 

# clean_coral %>% 
#   filter(taxa == "Por") %>% 
#   arrange(desc(volume)) %>% 
#   select(coral_number, year, length, width, height, volume) %>% 
#   head()
```

# Source volume functions

```{r}
source(here::here("R", "sph_vol.R"))

source(here::here("R", "hemisph_vol.R"))

source(here::here("R", "cyl_vol.R"))
```

# Calculate volume and lagged sizes

Removed rowwise() and the 3-way filter/rbind — replaced with a single case_when in mutate Dropped size_t0 = volume (redundant) Added ungroup() after the lagged size calculation Switched case_when to if_else for the single taxa recode here::here() on both write_csv calls Collapsed the initial cleaning into one piped chain

```{r}
volume_coral <- clean_coral %>%
  mutate(size_t0 = case_when(
    taxa == "Poc" ~ sph_vol(length, width, height),
    taxa == "Por" ~ hemisph_vol(length, width, height),
    TRUE          ~ cyl_vol(length, width, height)
  )) %>%
  arrange(coral_number, year) %>%
  group_by(coral_number) %>%
  mutate(
    size_t_minus1 = lag(size_t0),
    size_t1       = lead(size_t0)
  ) %>%
  ungroup()

write_csv(clean_coral,  here::here("data", "clean_corals.csv"))
write_csv(volume_coral, here::here("data", "volume_corals.csv"))
```

```{r}
ggplot(volume_coral, aes(x = taxa, y = size_t0, fill = taxa)) +
  geom_boxplot() +
  labs(x = "Taxa", y = "Log Volume (cm³)") +
  scale_y_log10(labels = scales::label_log()) + 
  scale_fill_manual(values = coral_colors) + 
  theme_minimal() + 
  theme(legend.position = "none")
```

```{r}
library(ggridges)


ggplot(volume_coral, aes(x = size_t0, y = taxa, fill = taxa)) +
  geom_density_ridges() +
  scale_x_log10(labels = scales::label_log()) +
  labs(x = "Log Volume (cm³)", y = "Taxa") +
  scale_fill_manual(values = coral_colors) + 
  theme_minimal() + 
    theme(legend.position = "none")
```
