---
title: "Coral Demography Data Cleaning"
format: html
editor: visual
authors: Kylie Newcomer, Joaquin Sandoval, Vedika Shirtekar
---

This document handles cleaning and quality assurance of long-format coral demography data spanning 2013–2023, as well as the standardization of 2024 field sheets to align with the long-format structure. Each row represents a single coral observation in a given year, with associated size measurements (length, width, height) and demographic dynamics (recruitment, death, fusion, fission etc). The cleaning steps standardize inconsistent labels in the `Dynamics` column, split it into binary indicator columns, and address logical inconsistencies in recruitment and death events to ensure each represents a true demographic transition rather than a carried-forward status.

### Load Necessary Libraries

```{r, message= FALSE}
library(tidyverse)
library(readxl)
library(here)
library(janitor)

```

### Import `Coral_Demography_allyears_2013-24_V1.xlsx` and 2024 raw field sheets

```{r,message = FALSE, warning=FALSE}
# Read in observations from 2013-2023 

wide_data_2013_2023 <- read_excel(here("rawdata", "Coral_Deomography_allyears_2013-24_V1.xlsx"), sheet = 1)

# Produce a character vector of the names of files or in 2024 sheets folder 

individual_files <- list.files("rawdata/2024sheets", pattern = "*.xlsx", full.names = TRUE)

```

### Combine 2024 field sheets into a single dataframe

```{r, message = FALSE, warning= FALSE}
# Function to standardize raw 2024 sheets with inconsistent colulmn names/ pull observer from first row above all columns 

read_file <- function(f) {
  meta <- read_excel(f, col_names = FALSE, n_max = 1)
  obs <- tail(na.omit(as.character(meta[1, ])), 1)
  read_excel(f, skip = 1, col_names = TRUE, col_types = "text") %>%
    mutate(Observer = obs)
}

# Combine into one dataframe
df_2024_sheets <- bind_rows(lapply(individual_files, read_file)) %>%
  # Deselect columns with previous year measurements 
    select(-c("W", "H", "L", "Status"))  %>%
  # Correct naming convention
  rename(habitat = Hab,
    transect = Tran) %>%
  # Standardize column naming structure across sheets 
  mutate(l = coalesce(L2024, L24),
         w = coalesce(W2024, W24),
         h = coalesce(H2024, H24),
  # Standardize dynamics column across sheets  
         dynamics = coalesce(Dynam, Dynamic, Dyn, Dynamics),
  # Convert coordinate and measurement columns to numeric type 
         across(c(X, Y, Z, l, w, h), as.numeric)
  ) %>%
  # Deselect irrelevent columns 
  select(-any_of(c("L2024", "W2024", "H2024", "L24", "W24", "H24",
                   "Dynam", "Dynamic", "Dyn", "Dynamics", "...17",
                   "status", "source_file"))) %>%
  # Add year column in which all rows are assigned 2024 
  mutate(year = 2024) %>% 
  # Standardize to lower-snake case 
  clean_names() %>% 
  # Drop rows where key measurement columns are all NA: caused by spaces in 2024 sheets 
  filter(!if_all(c(site, habitat, transect, taxa, l, w, h), is.na)) 
```

### Pivot 2013–2023 data into long format

```{r}
# Pivot 2013-2023 observations into long format 
long_data_2013_2023 <- wide_data_2013_2023 %>%
  # Deselect empty 2024 columns. These will be added from df_20204_sheets
  select(!contains("24")) %>% 
  pivot_longer(cols = -c(1:8), 
               names_to = c(".value", "year"), 
               names_pattern = "(.*)(\\d{4}$)") %>% 
  # Get rid of rows in which year is NA 
  filter(!is.na(year))
```

This is the long-format version of `wide_data_2013_2023`, where each coral observation in each year occupies its own row. The `Dynamics` column contains inconsistencies and needs to be split into separate columns, with `1` indicating the presence of a dynamic and `0` its absence.

### Split dynamics column into binary indicators

```{r}
# Add dynamics
long_data_2013_2023_with_dynamics <- long_data_2013_2023 %>% 
  # Takes a string and splits it into multiple rows
  separate_longer_delim(dynamics_, delim = regex("\\s*,\\s*")) %>% # regex for inconsistencies in spacing 
  # Create present indicator column 
  mutate(present = case_when(
    is.na(dynamics_) ~ 0, # If dynamics column is NA, present = 0 
    TRUE ~ 1),            # Otherwise, present = 1 
    # Standardizing the dynamics_ labels 
    dynamics_ = case_when(
                str_detect(dynamics_, "Fus") ~ "Fusion",
                str_detect(dynamics_, "Fis") ~ "Fission",
                TRUE ~ dynamics_ )) %>% 
  pivot_wider(names_from = dynamics_, values_from = present, names_prefix = "dyn_", values_fill = 0) %>% 
  clean_names() %>% 
  # Select relevant columns and order: dropping dyn_na
  select(coral_number, year, site, habitat, transect, taxa, x, y, z, l, w, h, obs, note, growth_ratio, dyn_recruitment, dyn_growth, dyn_shrinkage, dyn_death, dyn_immigration, dyn_emigration, dyn_fusion, dyn_fission, dyn_edge, dyn_missing_data)
```

Each unique coral should only have one recruitment and death event throughout the dataset. To regulate inconsistencies, we will isolate the corals and fix.

### Identify corals with multiple recruitment events

```{r}
# Run a test to see how many corals "recruited twice"

recruit_check <- long_data_2013_2023_with_dynamics %>%
  group_by(coral_number) %>%
  summarize(recruit_occurrences = sum(dyn_recruitment)) %>%
  filter(recruit_occurrences > 1)

nrow(recruit_check)
```

There are 192 corals with multiple recruitment events. Not ecologically plausible.

### Inspect multi-recruit coral histories

```{r}
# Look into each coral number in this list :

numerous_recruits <- long_data_2013_2023_with_dynamics %>%
  filter(coral_number %in% recruit_check$coral_number) %>% 
  arrange(coral_number, year) %>% 
   select(coral_number, year, l, w, h, dyn_recruitment, dyn_death) 


```

### Temporary fix: some have 3 recruitment events. May need to inspect further:

-   **Recruit → death → recruit**: Same `coral_number` reused after death. These are *two different corals.* Fix: assign new IDs to the second life, or drop.

-   **Recruit → recruit (no death between)** — Likely a labeling error; the second "recruit" is just continued growth. Fix: keep first, zero the rest

```{r}
long_data_2013_2023_with_dynamics <- long_data_2013_2023_with_dynamics %>% 
  arrange(coral_number, year) %>% 
  group_by(coral_number) %>% 
  mutate(dyn_recruitment = if_else(cumsum(dyn_recruitment) > 1, 0, dyn_recruitment), 
         year = as.numeric(year))

# Re-run test to verify fix worked
recruit_check <- long_data_2013_2023_with_dynamics %>%
  group_by(coral_number) %>%
  summarize(recruit_occurences = sum(dyn_recruitment)) %>%
  filter(recruit_occurences > 1)

nrow(recruit_check)

```

```{r}
death_test <- long_data_2013_2023_with_dynamics %>% 
  group_by(coral_number) %>% 
  mutate(death = sum(dyn_death)) %>% 
  filter(death > 1) %>% 
  select(coral_number, year, l, w, h, dyn_death)


```

```{r}

coral_test <- long_data_2013_2023_with_dynamics %>%
  mutate(recruit_year = case_when(
    dyn_recruitment == 1 ~ year,
    TRUE ~ NA),
     transect = case_when(coral_number %in% c(1347: 1353) ~ "T01",
                          coral_number %in% c(1361: 1363) ~ "T01", 
                          coral_number %in% c(1366: 1376) ~ "T01", 
                          coral_number %in% c(1466: 1483) ~ "T01",
                          coral_number == 1863 ~ "T04",
                          coral_number == 1872 ~ "T04", 
                          TRUE ~ transect)
  ) %>% 
  group_by(coral_number) %>% 
  mutate(recruit_year = sum(recruit_year, na.rm = TRUE)) %>% 
  mutate(recruit_year = case_when(
    recruit_year == 0 ~ 2013,
    TRUE ~ recruit_year
  ),
  coral_id = str_c(site, habitat, transect, x, y, z, sep = "_"), 
  transect_id = str_c(site, habitat, transect, sep = "_"),
  l = as.numeric(l),
  w = as.numeric(w),
  h = as.numeric(h))

df_coral_id <- df_2024_sheets %>% 
  mutate(coral_id = str_c(site, habitat, transect, x, y, z, sep = "_"))

coral_test %>% 
  filter(coral_number == 1247)


corals_all <- rbind(coral_test, df_coral_id)
 
 
```

```{r}
long_data_t <- long_data_2013_2023 %>% 
  filter(str_detect(transect, "T"))
```
